QPLIB: A Library of Quadratic Programming Instances

### 2017-02-01 QPLIB-1.0.tex
cleanup problem notation
- LPs and NLPs are always continuous
- LP/NLP + integer variables is called MILP/MINLP
- QP can be continuous and discrete
- closes #1

However, this diversity also implies that QP means very different things to different researchers. This is illustrated by the simple fact that the class of problems that we simply refer to here as ``QP'' is called in different ways, among which Mixed-Integer Quadratically Constrained Quadratic Problems (MI-QCQP). It is perhaps therefore not surprising that, unlike for ``simpler'' problems classes \cite{Koch2011}, so far there has never been a single library containing all different kinds of instances of QP. Several libraries devoted to special cases of QP are indeed available; however, each of them is either focussed on one application (a specific problem that can be modeled as QP), or on QPs with specific structural properties that make them suitable to be solved with some given class of algorithmic approaches. To the best of our knowledge, collecting a set of QP instances that is at the same time not overwhelmingly numerous and significant for the many different interested communities has not been attempted, yet. This work constitutes a first step in this direction.

@@

Optimization problems with quadratic constraints and/or objective function (QP) have been the subject of a considerable amount of research over the last 60 years. At least some of the rationale for this interest is likely due to the fact that they are the ``least-nonlinear nonlinear problems''. Hence, in particular for the convex case, tools and techniques that have been honed during decades of research for Linear Programming (LP), typically with integrality constraints (MILP), can often be extended to the quadratic case with at least less effort than what would be required for tackling general Non Linear Programming (NLP) problems, without or with integrality constraints (MINLP). This has indeed happened over and over again with different algorithmic techniques, such as Interior Point methods, active-set methods (of which the simplex method is a prototypical example), enumeration methods, cut-generation techniques, reformulation techniques, and many others \cite{BS09}.
Similarly, nonconvex continuous QP are perhaps the ``simplest'' class of problems that require features like spatial enumeration techniques to be solved.
Hence, they are both the natural basis for the development of general techniques for nonconvex NLP, and a very specific class so that specialized approaches can be developed \cite{Dur2010,Burer2012}.

On the other hand, (MI)QP is, in some sense, a considerably more expressive class than (MI)LP. Quadratic expressions are found, either naturally or after appropriate reformulations, in very many optimization problems \cite{Kochenberger2014}. Table \ref{tbl:application} provides a certainly non-exhaustive collection of applications that lead to formulations with either quadratic constraints, or quadratic objective function, or both. Also, in general any continuous function can be approximated with arbitrary accuracy (over a compact set) by a polynomial of arbitrary degree; in turn, every polynomial can be broken down to a system of quadratic expressions. Hence, (MI)QP is, in some sense, roughly as expressive as the whole of (MI)NLP. Of course this is, in principle, true for (MI)LP as well, but at the cost of much larger and much more complex formulations. Hence, for many applications QP may represent the ``sweet spot'' between the effectiveness, but lower expressive power, of (MI)LP and the higher expressive power, but much higher computational cost, of (MI)NLP.

@@

\caption{Application Domains of (MI)QP}\label{tbl:application} \\

@@

\multicolumn{3}{l}{\emph{Table \ref{tbl:application} (Application Domains of (MI)QP) continued}} \\

>>>>

However, this diversity also implies that QP means very different things to different researchers. This is illustrated by the simple fact that the class of problems that we simply refer to here as ``QP'' is called in different ways, among which Mixed-Integer Quadratically Constrained Quadratic Problems (MIQCQP). It is perhaps therefore not surprising that, unlike for ``simpler'' problems classes \cite{Koch2011}, so far there has never been a single library containing all different kinds of instances of QP. Several libraries devoted to special cases of QP are indeed available; however, each of them is either focussed on one application (a specific problem that can be modeled as QP), or on QPs with specific structural properties that make them suitable to be solved with some given class of algorithmic approaches. To the best of our knowledge, collecting a set of QP instances that is at the same time not overwhelmingly numerous and significant for the many different interested communities has not been attempted, yet. This work constitutes a first step in this direction.

@@

Optimization problems with quadratic constraints and/or objective function (QP) have been the subject of a considerable amount of research over the last 60 years. At least some of the rationale for this interest is likely due to the fact that they are the ``least-nonlinear nonlinear problems''. Hence, in particular for the convex case, tools and techniques that have been honed during decades of research for Linear Programming (LP), typically with integrality constraints (MILP), can often be extended to the quadratic case with at least less effort than what would be required for tackling general (Mixed-Integer) Nonlinear Programming ((MI)NLP) problems. This has indeed happened over and over again with different algorithmic techniques, such as Interior Point methods, active-set methods (of which the simplex method is a prototypical example), enumeration methods, cut-generation techniques, reformulation techniques, and many others \cite{BS09}.
Similarly, nonconvex continuous QP are perhaps the ``simplest'' class of problems that require features like spatial enumeration techniques to be solved.
Hence, they are both the natural basis for the development of general techniques for nonconvex NLP, and a very specific class so that specialized approaches can be developed \cite{Dur2010,Burer2012}.

On the other hand, QP, that is, quadratic problems with continuous or integer variables, is, in some sense, a considerably more expressive class than (MI)LP. Quadratic expressions are found, either naturally or after appropriate reformulations, in very many optimization problems \cite{Kochenberger2014}. Table \ref{tbl:application} provides a certainly non-exhaustive collection of applications that lead to formulations with either quadratic constraints, or quadratic objective function, or both. Also, in general any continuous function can be approximated with arbitrary accuracy (over a compact set) by a polynomial of arbitrary degree; in turn, every polynomial can be broken down to a system of quadratic expressions. Hence, QP is, in some sense, roughly as expressive as the whole of MINLP. Of course this is, in principle, true for MILP as well, but at the cost of much larger and much more complex formulations. Hence, for many applications QP may represent the ``sweet spot'' between the effectiveness, but lower expressive power, of MILP and the higher expressive power, but much higher computational cost, of MINLP.

@@

\caption{Application Domains of QP}\label{tbl:application} \\

@@

\multicolumn{3}{l}{\emph{Table \ref{tbl:application} (Application Domains of QP) continued}} \\

### 2017-02-01 methods.tex
cleanup problem notation
- LPs and NLPs are always continuous
- LP/NLP + integer variables is called MILP/MINLP
- QP can be continuous and discrete
- closes #1

Incomplete methods are usually realized as local search algorithms, asymptotically complete methods are usually realized by meta-heuristic methods such as multi-start or simulated annealing, and complete methods for $\mathcal{NP}$-hard problems such as QP are typically realized as implicit exhaustive exploration algorithms. However, these three categories may exhibit some overlap. For example, any deterministic method for solving QCQ locally is incomplete in general, but becomes complete for CCC, since any local optimum of a convex QP is also global. Therefore, when we state that a given algorithm is incomplete or (asymptotically) complete we mean that it is so the largest problem class that the solver naturally targets, although it may be complete on specific sub-classes. For example, interior point algorithms naturally target continuous NLPs and are incomplete on nonconvex NLPs, and therefore on QCQ, but become complete for CCC. In general, all complete methods for a problem $P$ must be complete for any problem $Q \subseteq P$, while a complete method for $P$ might be incomplete for $R \supset P$.

@@

The solvers in Table \ref{t:solvers} which implement incomplete methods for continuous NLPs (a problem class containing QCQ) are \conopt, \ipopt, \minos, \snopt, and \knitro. Note that all these solvers tackle the more general class of continuous NLP, while we use them only for the considerably more restricted class of QP. Aside from solvers provided by GAMS, there are a number of other, specialized incomplete QP solvers, such as
\cqp \cite{GoulOrbaRobi13:mpc},
\dqp \cite{GoulRobi16:coap} and
\ooqp \cite{GertWrig03:toms}
for convex problems, and
\bqpd \cite{Flet:2000},
\qpa \cite{GoulToin02i} and
\qpb \cite{ConnGoulOrbaToin:2000},
\qpc \cite{GoulOrbaToin03:toms},
\sqic \cite{Gill2015}
for nonconvex ones
While there are many different approaches for continuous NLPs in general, we mention here those that are particularly well-suited for QP, i.e.:
%
\begin{itemize}
 \item active set methods (\conopt, \minos, \snopt, \knitro, \bqpd, \qpa, \sqic) \cite{Dost09};
 %
 \item interior point methods (\ipopt, \knitro, \cqp, \ooqp, \qpb, \qpc) \cite{Wright97};
 %
 \item projected gradient methods (\dqp) \cite{CalaMore87:mp,ChenGui13:coap}.
\end{itemize}
%
Active set and interior point methods have been defined for *C* but also for general (continuous) NLPs, and many of the solvers named above target this
larger problem class.

@@

In general, the applicability of meta-heuristics to a given problem relies on availability of a local search that exploit the structural properties of the problem.
Depending on the local search employed, Multi-start methods can address MINLP of the most general class.

>>>>

Incomplete methods are usually realized as local search algorithms, asymptotically complete methods are usually realized by meta-heuristic methods such as multi-start or simulated annealing, and complete methods for $\mathcal{NP}$-hard problems such as QP are typically realized as implicit exhaustive exploration algorithms. However, these three categories may exhibit some overlap. For example, any deterministic method for solving QCQ locally is incomplete in general, but becomes complete for CCC, since any local optimum of a convex QP is also global. Therefore, when we state that a given algorithm is incomplete or (asymptotically) complete we mean that it is so the largest problem class that the solver naturally targets, although it may be complete on specific sub-classes. For example, interior point algorithms naturally target NLPs and are incomplete on NLPs, and therefore on QCQ, but become complete for CCC. In general, all complete methods for a problem $P$ must be complete for any problem $Q \subseteq P$, while a complete method for $P$ might be incomplete for $R \supset P$.

@@

The solvers in Table \ref{t:solvers} which implement incomplete methods for NLPs (a problem class containing QCQ) are \conopt, \ipopt, \minos, \snopt, and \knitro. Note that all these solvers tackle the more general class of NLP, while we use them only for the considerably more restricted class of QP. Aside from solvers provided by GAMS, there are a number of other, specialized incomplete QP solvers, such as
\cqp \cite{GoulOrbaRobi13:mpc},
\dqp \cite{GoulRobi16:coap} and
\ooqp \cite{GertWrig03:toms}
for convex problems, and
\bqpd \cite{Flet:2000},
\qpa \cite{GoulToin02i} and
\qpb \cite{ConnGoulOrbaToin:2000},
\qpc \cite{GoulOrbaToin03:toms},
\sqic \cite{Gill2015}
for nonconvex ones
While there are many different approaches for NLPs in general, we mention here those that are particularly well-suited for QP, i.e.:
%
\begin{itemize}
 \item active set methods (\conopt, \minos, \snopt, \knitro, \bqpd, \qpa, \sqic) \cite{Dost09};
 %
 \item interior point methods (\ipopt, \knitro, \cqp, \ooqp, \qpb, \qpc) \cite{Wright97};
 %
 \item projected gradient methods (\dqp) \cite{CalaMore87:mp,ChenGui13:coap}.
\end{itemize}
%
Active set and interior point methods have been defined for *C* but also for general NLPs, and many of the solvers named above target this
larger problem class.

@@

In general, the applicability of meta-heuristics to a given problem relies on availability of a local search that exploit the structural properties of the problem.
Depending on the local search employed, multi-start methods can address MINLP of the most general class.
